Determinism, Complexity, and Predictability in Computer Performance

### 2014-08-04 model.tex
Liz final read-through round II

These four forecast methods were chosen to be a representative
sampling of standard prediction strategies, but they do not, of
course, cover that space exhaustively.  Our goal here is an empirical
assessment of the relationship between predictability and complexity,
not formal results about a ``best'' predictor for a given time series.
From a practitioner's standpoint, it would be useful to know \emph{ a
  priori} if a time series contains enough predictive structure to
make it worth spending the time and effort searching for a good
forecast method.  It would also be useful to know if a given method is
inadequate---that is, if an alternative method could do better and so
one should continue searching.  \alert{Liz to do: check that this is
  not redundant, given the other text in this section (and elsewhere
  in the paper).  it may be better to put this in the conclusion.}

>>>>

% These four forecast methods were chosen to be a representative
% sampling of standard prediction strategies, but they do not, of
% course, cover that space exhaustively.  Our goal here is an empirical
% assessment of the relationship between predictability and complexity,
% not formal results about a ``best'' predictor for a given time series.
% From a practitioner's standpoint, it would be useful to know \emph{ a
%   priori} if a time series contains enough predictive structure to
% make it worth spending the time and effort searching for a good
% forecast method.  It would also be useful to know if a given method is
% inadequate---that is, if an alternative method could do better and so
% one should continue searching.  \alert{Liz to do: check that this is
%   not redundant, given the other text in this section (and elsewhere
%   in the paper).  it may be better to put this in the conclusion.}

### 2014-08-04 results.tex
Liz final read-through round II

The discussion in the previous paragraph highlights the second finding
of this paper: the ability of our proposed heuristic to flag
inappropriate models.  \arima is an automated, mechanical procedure
for choosing parameters for an ARIMA model of a given data set.  While
the tests and criteria employed by this algorithm
(Section~\ref{sec:arima}) are sophisticated, the results can still be
sub-optimal---if the initial space of models being searched is not
broad enough, for instance, or if one of the preliminary tests gives
an erroneous result.  \arima \emph{always} returns a model, and it can
be very hard to detect when that model is bad.  Our results suggest a
way to do so: if the MASE score of an auto-fitted model like \arima is
out of line with the WPE value of the data, that can be an indication
of inappropriateness in the order selection and parameter estimation
procedure.  A close examination of the \arima MASE scores and WPE
values of the 15 \col traces, for instance, brought out such a
situation: all of these signals had the same complexity, but some (the
outliers in the blue cluster of {\color{blue}$+$} points in the
bottom-left panel of Figure~\ref{fig:wpe_vs_mase_all}) had much higher
MASE values, indicating a failure in the \arima process.  We then went
back to the data, noticed some shifts of the baseline in those
signals, and manually assisted the automated fitting procedure to
force the inclusion of a term that is appropriate for a nonstationary
signal.  In all cases, this produced ARIMA models with MASE scores in
line with the others in that cluster.

@@

Again, the curves in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_all} were determined from a finite set of
methods and data.  We put a lot of thought and effort into making
these experiments upon which these results are based both
representative and comprehensive.  The forecast methods involved range
from the simple to the sophisticated; the time-series data was sampled
from a system whose behavior spans the dynamical behavior space.
While we are cautiously optimistic about the generality of our
conclusions, more exploration will be required before we can make any
such claim.  Our preliminary work along those lines shows that data
from the Henon map \cite{henon}, the Lorenz system \cite{lorenz}, and
the SFI A data set \cite{weigend-book,sfi-data} all fall within the
one-$\sigma$ volume of the fit in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_all} region, as do various nonlinear
transformations of \svdtwo, \svdfive and \svdsix.

>>>>

The discussion in the previous paragraph highlights the second finding
of this paper: the ability of our proposed heuristic to flag
inappropriate models.  \arima is an automated, mechanical procedure
for choosing parameters for an ARIMA model of a given data set.  While
the tests and criteria employed by this algorithm
(Section~\ref{sec:arima}) are sophisticated, the results can still be
sub-optimal---if the initial space of models being searched is not
broad enough, for instance, or if one of the preliminary tests gives
an erroneous result.  \arima \emph{always} returns a model, and it can
be very hard to detect when that model is bad.  Our results suggest a
way to do so: if the MASE score of an auto-fitted model like \arima is
out of line with the WPE value of the data, that can be an indication
of inappropriateness in the order selection and parameter estimation
procedure.

% A close examination of the \arima MASE scores and WPE values of the 15
% \col traces, for instance, brought out such a situation: all of these
% signals had the same complexity, but some (the outliers in the blue
% cluster of {\color{blue}$+$} points in the bottom-left panel of
% Figure~\ref{fig:wpe_vs_mase_all}) had much higher MASE values,
% indicating a failure in the \arima process.  We then went back to the
% data, noticed some shifts of the baseline in those signals, and
% manually assisted the automated fitting procedure to force the
% inclusion of a term that is appropriate for a nonstationary signal.
% In all cases, this produced ARIMA models with MASE scores in line with
% the others in that cluster.

@@

The curves in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_all} were determined from finite sets of
methods and data.  We put a lot of thought and effort into making
these sets representative and comprehensive.  The forecast methods
involved range from the simple to the sophisticated; the time-series
data analyzed in this section was sampled from a system whose behavior
spans the dynamical behavior space.  While we are cautiously
optimistic about the generality of our conclusions, more exploration
will be required before we can make any such claim.  Our preliminary
work along those lines shows that data from the Henon map
\cite{henon}, the Lorenz system \cite{lorenz}, the SFI A data set
\cite{weigend-book,sfi-data}, and a white-noise process all fall
within the one-$\sigma$ volume of the fit in
Figures~\ref{fig:wpe_vs_mase_best} and~\ref{fig:wpe_vs_mase_all}
region, as do various nonlinear transformations of \svdtwo, \svdfive
and \svdsix.
